Journey: A Study of Studying Large Memory Systems (TODO: should we get a new title?)

### 2017-04-20 paper.tex
Updated Intro, Related Work; added Backgrnd section

\usepackage[margin=1in]{geometry}
\usepackage{caption}

\title{Journey: A Study of Studying Large Memory Systems}
\author{Mark Mansi, Suhas Pai, Hasnain Ali Pirzada}
\date{}

@@

Currently, there are few good tools available to systems researchers for
understanding how systems interact with large memories. The ideal end-goal of
this research is to identify bottlenecks in managing large memories and find a
general lightweight technique or tool for studying and designing operating
systems capable of handling large memory while using systems currently
available to researchers.

(TODO: we might want to add stuff about using swap instead of larger memory)

@@

@@

The emergence of the internet required businesses to buy expensive machines to
keep up with demand for their services, but systems researchers did not have
access to these machines. Alameldeen et al. describe how they simulate a
multi-million dollar server using a \$2000 workstation. To do this, they go
through several rounds of scaling down, optimizing, changing, and tuning both
the benchmarks and systems they test. For example, they scale down a workload to
fit in the 1GB memory of the machine they use and increase the number of threads
to improve parallelism \cite{2kmachine}. This methodology may be accurate, but
it requires modifying the benchmarks, which is error prone and time consuming;
it can be extremely difficult to validate changes to a benchmark. In our work,
we seek to avoid changes to the experimental workloads including changes to code
or scale.

Likewise, the Quartz emulator studies the behaviour of applications on
Non-volatile Memory (NVM) while actually running on top of DRAM. It uses
existing hardware capabilities to emulate the higher latency and lower bandwidth
of most NVM technologies. While Quartz does not address the fact that the actual
memory available to the applications on NVM might be orders of magnitude higher
than on DRAM, it demonstrates that new technologies may be emulated using
existing technology \cite{quartz}.

The Simics simulator uses demand paging for simulated environments, which means
that memory is only allocated when it is used. As long as the working set of the
target system fits in the host memory, performance will be tolerable.  However,
in our work, we wish to study OS behavior under workloads that actually use all
available memory. Thus, the usefulness of demand paging is greatly reduced. For
these workloads, the performance of simulators like Simics tends to be
impractical \cite{simics}. We do not explore simulation further because it can
be orders of magnitude slower than native execution, making it infeasible for
studying large workloads or systems \cite{2kmachine}.

A similar problem was encountered by researchers studying both large storage
drives and large distributed storage systems. David is a system that allows
storage and big data researchers to run large benchmarks requiring terabytes of
storage using off the shelf storage devices (which at that time were too small).
David creates a compressed version of the file system by physically storing only
metadata and discarding the contents of files. Reads to the disk causes David to
generate data on the fly. This decision choice is based on the observation that
most benchmarking frameworks do not care about the actual content of the files,
and that most of the storage capacity of a drive tends to be data rather than
metadata \cite{david}. The Exalt system uses a similar methodology for
large-scale distributed storage systems \cite{exalt}. This methodology provides
a promising direction for large-memory system studies. One may consider ignoring
the contents of a process’s heap and only storing kernel data structures and a
process’s code and stack segments. However, generating heap contents on the fly
is more difficult than generating disk contents because of the common use of
custom data structures. Moreover, on Linux, the kernel data structures for 1TB
of memory may also exceed the size of physical memory on present systems
\cite{simics}.

A virtualized environment can be used to provide a guest system with more memory
than is physically available to the host. A study of the limits of the KVM
hypervisor found that there is no fundamental limit to the size of guest
physical memory other than the hardware address width. However, currently, a
Linux host running KVM will require the guest memory to be backed by host
physical memory or host swap space \cite{ibmkvm}.  This means that when the
virtual machine uses the whole amount of memory allocated to it, the host can
swap pages to disk. The resulting poor performance can cause inaccurate
performance measurements when running benchmarks.

Prefetching pages from swap space can offer a way to mitigate the overhead of
memory overcommitment. When there is significant memory pressure, even pages
which are likely to be accessed soon are swapped out to disk. They are faulted
back into memory when accessed, resulting in significant performance cost.
Charm++ uses a programming model where computations can be scheduled by the
language runtime. A designated thread can prefetch pages required by a
computation, averting page faults. Although this approach is effective in
avoiding overhead, it requires applications to be written with a particular
programming language and so is not applicable to most benchmarks without
changing them \cite{charmpp}.

Gupta et al. built an emulator for high speed networks using a technique called
time dilation, which slows down the OS clock to make it appear that external
events are occurring faster. This allows the system to emulate network links
with speeds that are currently not available. The implementation is based on the
VMs and Xen hypervisor; Xen delivers the timer interrupts to the guest at a
lower rate than hardware hence  slowing down the guest’s clock \cite{timedil}.
We may use a similar approach to slow down system time  while paging in and out
large portions of memory. This will allow us to make the system believe that it
is reading data from the memory while actually most of that data is being read
from the disk.

Finally, there have been studies which look at the performance overheads
associated with current implementations of virtual memory and suggest mechanisms
to mitigate them. RadixVM tries to overcome performance issues in highly
concurrent workloads due to serialization of memory management operations on
kernel data structures \cite{radixvm}. This work demonstrates that many parts of
existing memory management schemes are not scalable to larger systems.
Similarly, in our work, we wish to examine scalability limitations of memory
management in the Linux kernel. Other studies have suggested that \texttt{struct
page}, \texttt{struct vm\_area\_struct}, and page tables tend to comprise a
large portion of memory management overhead \cite{simics}.

@@

Our methodology is broken into two parts. First, we measure the scalability of
Linux's \texttt{vm\_area\_struct}. Then, we measure the minimum amount of memory
necessary for swapping.

@@

>>>>

\usepackage[margin=1in]{geometry}
\usepackage{caption}
\usepackage{graphicx}

\title{Journey: A Study of Studying Large Memory Systems (TODO: should we get a new title?)}
\author{Mark Mansi, Suhas Pai, Hasnain Ali Pirzada}
\date{}

@@

Currently, there are few good tools available to systems researchers for
understanding how systems interact with large memories. We see the need for a
general lightweight technique or tool for studying and designing operating
systems capable of handling large memory while using systems currently
available to researchers.  An important first step towards building such a tool
is to identify bottlenecks in managing large memories that make a tool
difficult to develop.

In particular, we look at the case of a single process using a large amount of
memory in Linux. We run experiments to identify the costs of
\texttt{vm\_area\_struct}, a common bookkeeping data structure in the Linux
kernel, along with the memory and processing overheads of swapping. From these
data, we extrapolate to systems running with more memory with an eye on
implications for studying large systems. While these experiments are not
comprehensive, we believe they can offer insights that are useful both for
building a research tool and for building systems.

@@

(TODO: do we need to refocus this section? or move parts of it to an appendix?)

@@

The emergence of the internet required businesses to buy expensive machines to
keep up with demand for their services, but systems researchers did not have
access to these machines. Alameldeen et al. describe how they simulate a
multi-million dollar server using a \$2000 workstation. To do this, they go
through several rounds of scaling down, optimizing, changing, and tuning both
the benchmarks and systems they test. For example, they scale down a workload to
fit in the 1GB memory of the machine they use and increase the number of threads
to improve parallelism \cite{2kmachine}. This methodology may be accurate, but
it requires modifying the benchmarks, which is error prone and time consuming;
it can be extremely difficult to validate changes to a benchmark. Ideally,
researchers can run large experimental workloads without changes, but the overheads
of current memory management systems prevent this from being practical.

Likewise, the Quartz emulator studies the behaviour of applications on
Non-volatile Memory (NVM) while actually running on top of DRAM. It uses
existing hardware capabilities to emulate the higher latency and lower bandwidth
of most NVM technologies. While Quartz does not address the fact that the actual
memory available to the applications on NVM might be orders of magnitude higher
than on DRAM, it demonstrates that new technologies may be emulated using
existing technology \cite{quartz}. (TODO: is this system still relavent?)

The Simics simulator uses demand paging for simulated environments, which means
that memory is only allocated when it is used. As long as the working set of
the target system fits in the host memory, performance will be tolerable.
However, in our work, we wish to study OS behavior under workloads that
actually use all available memory. Thus, the usefulness of demand paging is
greatly reduced because swapping causes disk latency to become the main
performance bottleneck. Moreover, the performance overhead of simulators like
Simics tends to be impractical because system state needs to be maintained by
the simulator \cite{simics}. We do not explore simulation further because it
can be orders of magnitude slower than native execution, making it infeasible
for studying large workloads or systems \cite{2kmachine}. However, the Simics
simulator illustrates the usefulness of memory management techniques such as
demand paging and the potential overheads of swapping.

A similar problem was encountered by researchers studying both large storage
drives and large distributed storage systems. David is a system that allows
storage and big data researchers to run large benchmarks requiring terabytes of
storage using off the shelf storage devices (which at that time were too small).
David creates a compressed version of the file system by physically storing only
metadata and discarding the contents of files. Reads to the disk causes David to
generate data on the fly. This decision choice is based on the observation that
most benchmarking frameworks do not care about the actual content of the files,
and that most of the storage capacity of a drive tends to be data rather than
metadata \cite{david}. The Exalt system uses a similar methodology for
large-scale distributed storage systems \cite{exalt}. This methodology provides
a promising direction for large-memory system studies. One may consider ignoring
the contents of a process’s heap and only storing kernel data structures and a
process’s code and stack segments. However, generating heap contents on the fly
is more difficult than generating disk contents because of the common use of
custom data structures. Moreover, on Linux, the kernel data structures for 1TB
of memory may also exceed the size of physical memory on current systems
\cite{simics}.

A virtualized environment can be used to provide a guest system with more memory
than is physically available to the host. A study of the limits of the KVM
hypervisor found that there is no fundamental limit to the size of guest
physical memory other than the hardware address width. However, currently, a
Linux host running KVM will require the guest memory to be backed by host
physical memory or host swap space \cite{ibmkvm}.  This means that when the
virtual machine uses the whole amount of memory allocated to it, the host can
swap pages to disk. The resulting poor performance can cause inaccurate
performance measurements when running benchmarks. It can also make large
benchmarks impractical.

Interestingly, many of non-volatile memory technologies are making their ways
into storage drives, such as Intel's recently-announced Optane SSDs. One
potential use of fast storage devices is for fast swapping space. Some
literature has proposed that small amounts of DRAM backed by large amounts of
fast swap space can be a scalable alternative to having large amounts of fast
RAM (TODO: cite). (TODO: can we have this in our report?). In such a system,
the overhead of swapping may become a bottleneck.

Prefetching pages from swap space can offer a way to mitigate the overhead of
memory overcommitment, possibly making large virtual machines a viable
mechanism for studying large memory systems. When there is significant memory
pressure, even pages which are likely to be accessed soon are swapped out to
disk. They are faulted back into memory when accessed, resulting in significant
performance cost.  Charm++ uses a programming model where computations can be
scheduled by the language runtime. A designated thread can prefetch pages
required by a computation, averting page faults \cite{charmpp}. Rather than
implement page prefetching in a language runtime, one might implement a
prefetcher in the swapping subsystem, making it general purpose. A large body
of work already exists on hardware prefetching for processor caches, on which a
swap prefetcher might draw for inspiration (TODO: cite). However, this approach
can only work if page access patterns are predictable. Moreover, the high
latency of disks implies that prefetchers would have to predict the very
distant future (e.g. seconds ahead of execution). (TODO: Mike's comment on the
previous version of this paragraph said that it was ``too distant/off topic''.
Is it better now? I feel like prefetching was an idea that we considered too
much to leave out...)

Gupta et al. built an emulator for high speed networks using a technique called
time dilation, which slows down the OS clock to make it appear that external
events are occurring faster. This allows the system to emulate network links
with speeds that are currently not available. The implementation is based on
the VMs and Xen hypervisor; Xen delivers the timer interrupts to the guest at a
lower rate than hardware hence  slowing down the guest’s clock \cite{timedil}.
A similar approach may be applied for large memory system studies to slow down
guest time while paging in and out large portions of memory. This will allows
the system to believe that it is reading data from the memory while actually
most of that data is being read from the disk.

Finally, there have been studies which look at the performance overheads
associated with current implementations of virtual memory and suggest mechanisms
to mitigate them. RadixVM tries to overcome performance issues in highly
concurrent workloads due to serialization of memory management operations on
kernel data structures \cite{radixvm}. This work demonstrates that many parts of
existing memory management schemes are not scalable to larger systems.
Similarly, in our work, we wish to examine scalability limitations of memory
management in the Linux kernel. Other studies have suggested that \texttt{struct
page}, \texttt{struct vm\_area\_struct}, and page tables tend to comprise a
large portion of memory management overhead \cite{simics} (TODO: cite the LWN
paper).

\section{Background}

In our methodology, we run several experiments to measure the scalability of
the Linux memory management system. Each experiment is designed to probe a
different aspect of the system. This section gives some background knowledge
necessary to understanding the experiments.

\subsection{Virtual Memory Areas}

A process may use the \texttt{mmap} system call to request that the kernel
allocate part of its virtual address space. Accesses to memory addresses
outside of allocated regions result in a segmentation fault. This makes memory
corruption errors easier to find and debug, and increases security from memory
corruption attacks.

The Linux kernel keeps track of which portions of a process's address space
have been allocated using the \texttt{vm\_area\_struct}. Because processes may
map a number of regions, including shared objects, stack, heap, text, and
anonymous regions, the scalability of this structure is critical to Linux's
performance.

When a process maps a region, a \texttt{vm\_area\_struct} is created, but no
memory is allocated. When a process first touches a page, a page fault occurs
and the processor traps into a page fault handler, which can allocate memory
and set up a page table entry. This is called demand paging.

For demand paging to work efficiently, the correct \texttt{vm\_area\_struct}
needs to be efficient to find. Linux caches the most recently used region and
organizes all regions into a balanced binary tree for fast lookup.  Moreover,
adjacent regions with similar permissions and properties are merged in an
effort to keep bookkeeping costs low.

\subsection{Swapping}

\subsection{Page Frame Reclamation}

@@

We conduct a series of experiments to measure the scalability of various parts of the Linux memory management system.
First, we measure the scalability of
Linux's \texttt{vm\_area\_struct}. Then, we measure the minimum amount of memory
necessary for swapping.

@@

\begin{figure}
    \includegraphics[width=\columnwidth]{memory_controled}
    \caption{$\Delta_i$ adjusted with control for benchmarks}
\end{figure}

\begin{figure}
    \includegraphics[width=\columnwidth]{vm_area_struct_count}
    \caption{Number of \texttt{vm\_area\_struct}s for benchmarks}
\end{figure}

TODO: memory results

TODO: time results
